---
title: "Gene Ontology Enrichment Analysis"
author:
- name: Kellen Cresswell
  affiliation:
  - &1 Department of Biostatistics, Virginia Commonwealth University, Richmond, VA
- name: Mikhail Dozmorov
  affiliation:
  - *1
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Gene Ontology Enrichment Analysis}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    chunk_output_type: console
bibliography: pack_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Using the output of `TADCompare` and `TimeCompare`, we can do a range of analyses. One common one is gene ontology enrichment analysis to determine the pathways in which genes near TAD boundaries occur in. To do this, we use [rGREAT](https://bioconductor.org/packages/release/bioc/html/rGREAT.html) an R package for performing gene ontology enrichment analysis. 

## Performing gene ontology analysis using TADCompare

In the first example, we show how to perform gene ontology enrichment using differential boundaries. Here, we perform the analysis on shifted boundaries detected in matrix 1.

```{r, warning=FALSE, message = FALSE}
library(rGREAT)
#Reading in data
data("rao_chr22_prim")
data("rao_chr22_rep")
#Performing differential analysis
results = TADCompare(rao_chr22_prim, rao_chr22_rep, resolution = 50000)
#Saving the results into its own data frame
TAD_Frame = results$TAD_Frame
#Filter data to only include complex boundaries enriched in the first contact matrix
TAD_Frame = TAD_Frame %>% dplyr::filter( (Type == "Shifted") & (Enriched_In == "Matrix 2")  )
#Assign a chromosome and convert to a bed format
TAD_Frame = TAD_Frame %>% dplyr::select(Boundary) %>%
  mutate(chr = "chr22", start = Boundary, end = Boundary) %>% dplyr::select(chr,start,end)
#Set up rGREAT job with default parameters
great_shift = submitGreatJob(TAD_Frame, request_interval = 1, version = "2.0")
#Submit the job
enrichment_table = getEnrichmentTables(great_shift)
#Subset to only include vital information
enrichment_table = bind_rows(enrichment_table, .id = 'source') %>% dplyr::select(Ontology = `source`, Description = `name`, `P-value` = Hyper_Raw_PValue)
#Print head organizaed by p-values
head(enrichment_table %>% dplyr::arrange( `P-value`))
```

The first column "Ontology" is simply the ontology which the pathway comes from. Here, we use the default which is GO ontologies but for more available ontologies see the [rGREAT vignette](https://bioconductor.org/packages/release/bioc/vignettes/rGREAT/inst/doc/rGREAT.html). "Description" is the pathway itself. "P-value" is the unadjusted hypergeometric p-value, as output by `rGREAT`. `rGREAT` also provides binomial p-values (Binom_Raw_Pvalue, Binom_Adjp_BH) and adjusted hypergeometric p-values (Hyper_Adjp_BH). 

Now we demonstrate how to perform the same analysis but for all boundary types simultaneously. In this case, we use time-varying data. 

```{r}
#Read in time course data
data("time_mats")
#Identifying  boundaries
results = TimeCompare(time_mats, resolution = 50000)
#Pulling out the frame of TADs
TAD_Frame = results$TAD_Bounds
#Getting sets of coordinates for TAD boundaries and converting into bed format
Bound_List = lapply(unique(TAD_Frame$Category), function(x) {
  TAD_Frame %>% filter((Category == x))  %>% mutate(chr = "chr22") %>% dplyr::select(chr, Coordinate) %>%
  mutate(start = Coordinate, end = Coordinate) %>% dplyr::select(chr,start,end)
})
#Performing rGREAT analysis for each boundary Category
TAD_Enrich = lapply(Bound_List, function(x) getEnrichmentTables(submitGreatJob(x, request_interval = 1, version = "2.0")))
#Name list of data frames to keep track of which enrichment belongs to which
names(TAD_Enrich) = unique(TAD_Frame$Category)
#Bind each category of pathway and create new column for each pathway
TAD_Enrich = lapply(names(TAD_Enrich), function(x) bind_rows(lapply(TAD_Enrich[[x]], function(y) y %>% mutate(Category = x)), .id = 'source')) 
#Bind each boundary category together and pull out important variables
enrichment_table = bind_rows(TAD_Enrich) %>% dplyr::select(Ontology = `source`, Description = `name`, `P-value` = Hyper_Raw_PValue, Category)
#Get the top enriched pathways
head(enrichment_table %>% dplyr::arrange(`P-value`))
```

These columns are the same as the differential analysis but with an extra column, "Category", indicating the type of time-varying TAD boundary. 
