---
title: "SpectralTAD Vignette"
author:
- name: Kellen Cresswell
  affiliation:
  - &1 Department of Biostatistics, Virginia Commonwealth University, Richmond, VA
- name: Mikhail Dozmorov
  affiliation:
  - *1

date: '`r format(Sys.Date(), "%B %e, %Y")`'
abstract: >
    TADCompare is an R package designed for identification and analysis of
    differential Topoloigcally Associated Domains (TADs) across multiple
    datasets. It uses a novel metric called boundary score that identifies
    the likelihood of TAD boundaries and can be used to quantify difference
    in TAD boundaries across datasets. We provide three unique functions:
    TADCompare for differential TAD boundary identification, TimeCompare
    for quantification of change over time and ConsensusTADs for consensus
    TAD calling. 
package: TADCompare
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{TADCompare}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    chunk_output_type: console
bibliography: pack_ref.bib
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
options(stringsAsFactors = FALSE)
```

# Introduction

`TADCompare` is an R package for analysis of differential TAD boundaries. It is designed to work on a wide-range of formats and resolutions of HiC data. `TADCompare` contains three functions: `TADCompare`, `TimeCompare` and `ConsensusTADs`. `TADCompare` allows for identification of differential TAD boundaries between two contact matrices. `TimeCompare` takes a set of contact matrices, one matrix per time point, identifies TAD boundaries and classifies how they change over time. `ConsensusTADs` takes a list of TADs and identifies a consensus of TAD boundaries across all matrices using our novel consensus boundary score. All three functions can take matrices in sparse 3-column format, $n \times n$, or $n \times (n+3)$ formats. [??? TBD the following] None of the functions have required parameters but instructions on how to alter them are provided. 

# Getting Started

## Installation

```{r, eval = FALSE, message=FALSE}
# if (!requireNamespace("BiocManager", quietly=TRUE))
#     install.packages("BiocManager")
# BiocManager::install("TADCompare")
devtools::install_github("cresswellkg/TADCompare")
library(TADCompare)
```

## Input data

### Working with $n \times n$ matrices

$n \times n$ contact matrices are most commonly associated with data coming from the Bing Ren lab (http://chromosome.sdsc.edu/mouse/hi-c/download.html). These contact matrices are square and symmetric with entry $ij$ corresponding to the number of contacts between region $i$ and region $j$. Below is an example of a $5 \times 5$ region of an $n \times n$ contact matrix derived from Rao et al. 2014 data, GM12878 cell line [@Rao:2014aa], chromosome 22, 50kb resolution. Note the symmetry around the diagonal - the typical shape of chromatin interaction matrix.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
data("rao_chr22_prim")
row.names(rao_chr22_prim) = colnames(rao_chr22_prim) = format(as.numeric(row.names(rao_chr22_prim)), scientific = FALSE)
coords <- 200:275
pheatmap::pheatmap(log10(rao_chr22_prim[coords, coords]), cluster_rows = FALSE, cluster_cols = FALSE)
```

### Working with $n \times (n+3)$ matrices

$n \times (n+3)$ matrices are commonly associated with the TopDom TAD caller (http://zhoulab.usc.edu/TopDom/). These matrices consist of a normal $n \times n$ matrix but with three additional leading columns containg the chromosome, the start of the region and the end of the region. Regions in this case are determined by the resolution of the data. The typical $n \times (n+3)$ matrix is shown below.

```{r, echo = FALSE, warning = FALSE}
sub_mat = data.frame(chr = "chr22", 
                     start = as.numeric(colnames(rao_chr22_prim[coords, coords])), 
                     end = as.numeric(colnames(rao_chr22_prim[coords, coords])) + 50000, 
                     rao_chr22_prim[coords, coords]) 
row.names(sub_mat) = NULL
sub_mat
```

### Working with sparse 3-column matrices

Sparse 3-column matrices are matrices where the first and second column refer to region $i$ and region $j$ of the chromosome, and the third column is the number of contacts between them.  This style is becoming increasingly popular and is associated with raw data from Lieberman-Aiden lab (e.g., https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525), and is the data output produced by the Juicer tool [@Durand:2016aa]. 3-column matrices are handled internally in the package by converting them to $n \times n$ matrices using the `HiCcompare` package's `sparse2full()` function. The first 5 rows of a typical sparse 3-column matrix is shown below.

```{r, echo = FALSE, warning = FALSE}
data("rao_chr22_prim") 
head(HiCcompare::full2sparse(rao_chr22_prim))
```

### Working with other data types

Users can also find TADs from data output by `cooler` (http://cooler.readthedocs.io/en/latest/index.html) and HiC-Pro (https://github.com/nservant/HiC-Pro) with minor pre-processing using the `HiCcompare` package.

#### Working with .cool files

The cooler software can be downloaded from https://mirnylab.github.io/cooler/. A catalog of popular HiC datasets can be found at ftp://cooler.csail.mit.edu/coolers. We can extract chromatin interaction data from .cool files using the following steps:

1. Follow instructions to install the cooler software, https://mirnylab.github.io/cooler/
[??? Does not work, IOError: Unable to open file (unable to open file: name = 'ftp://cooler.csail.mit.edu/coolers/hg19/Zuin2014-HEK293CtcfControl-HindIII-allreps-filtered.50kb.cool', errno = 2, error message = 'No such file or directory', flags = 0, o_flags = 0)]
2. Convert first matrix to text file using `cooler dump --join ftp://cooler.csail.mit.edu/coolers/hg19/Zuin2014-HEK293CtcfControl-HindIII-allreps-filtered.50kb.cool > Zuin.HEK293.50kb.Control.txt`
3. Convert first matrix to text file using `cooler dump --join ftp://cooler.csail.mit.edu/coolers/hg19/Zuin2014-HEK293CtcfDepleted-HindIII-allreps-filtered.50kb.cool > Zuin.HEK293.50kb.Depleted.txt`
4. Run the code below

```{r, eval = FALSE}
#Read in data
cool_mat1 = read.table("Zuin.HEK293.50kb.Control.txt")
cool_mat2 = read.table("Zuin.HEK293.50kb.Depleted.txt")

#Convert to sparse 3-column matrix using cooler2sparse from HiCcompare
sparse_mat1 = HiCcompare::cooler2sparse(cool_mat1)
sparse_mat2 = HiCcompare::cooler2sparse(cool_mat2)

#Run TADCompare
diff_tads = lapply(names(sparse_mat1), function(x) {
  TADCompare(sparse_mat1[[x]], sparse_mat2[[x]], resolution = 50000)
})

```

#### Working with HiC-Pro files

HiC-Pro data is represented as two files, the `.matrix` file and the `.bed` file. The `.bed` file contains a four columns (chromosome, start, end, ID). The `.matrix` file is a three-column matrix where the 1^st^ and 2^nd^ columns contain region IDs which map back to the coordinates in the bed file and the third column contains the number of contacts between the two regions. The steps for analyzing these files is shown below:

```{r, eval = FALSE}
#Read in both files
[??? You should include these data in the package. Select better data, describe where it came from]
mat1 = read.table("amyg_100000.matrix")
bed1 = read.table("amyg_100000_abs.bed")

#Matrix 2

mat2 = read.table("dplfc_100000.matrix")
bed2 = read.table("dplfc_100000_abs.bed")

#Convert to modified bed format
sparse_mats1 = HiCcompare::hicpro2bedpe(mat1,bed1)
sparse_mats2 = HiCcompare::hicpro2bedpe(mat2,bed2)

#Remove empty matrices if necessary
#sparse_mats$cis = sparse_mats$cis[sapply(sparse_mats, nrow) != 0]


#Go through all pairwise chromosomes and run TADCompare
sparse_tads = lapply(1:length(sparse_mats1$cis), function(z) {
  x = sparse_mats1$cis[[z]]
  y = sparse_mats2$cis[[z]]
  
  #Pull out chromosome
  chr = x[,1][1]
  #Subset to make three column matrix
  x = x[,c(2,5,7)]
  y = y[,c(2,5,7)]
  #Run SpectralTAD
  comp = TADCompare(x,y, resolution = 100000)
  return(list(comp, chr))
})

#Pull out differential TAD results
diff_res = lapply(sparse_tads, function(x) x$comp)
#Pull out chromosomes
chr = lapply(sparse_tads, function(x) x$chr)
#Name list by corresponding chr
names(diff_res) = chr
```

## Running TADcompare

The only required input is two contact matrices. TADCompare will automatically determine the type of matrix and convert it to an appropriate form, given it is one of the supported formats. The only requirement is that all matrices be in the same format. For fastests results, we suggest using $n \times n$ matrices. Additionally, we suggest users to provide resolution of their data. If resolution is not provided, we estimate it using the numeric column names of contact matrices. 

```{r, message = FALSE, warning = FALSE}
#Get the rao contact matrices built into the package
data("rao_chr22_prim")
data("rao_chr22_rep")
#We see these are n x n matrices
dim(rao_chr22_prim)
dim(rao_chr22_rep)
#Running the algorithm with resolution specified
results = TADCompare(rao_chr22_prim, rao_chr22_rep, resolution = 50000)
#Printing the first 5 differential boundaries
head(results$Diff_Loci, 5)
[??? Need to describe the structure of the results object. It is involved, so should be a separate paragraph]
#Repeating without specifying resolution
no_res = TADCompare(rao_chr22_prim, rao_chr22_rep)
#We can see below that resolution can be estimated automatically if necessary
identical(results$Diff_Loci, no_res$Diff_Loci)
```

## Running TimeCompare

`TimeCompare` takes a list of matrices similar in format to `TADCompare`. Like `TADCompare`, `TimeCompare` will estimate resolution and convert matrices to the appropriate format. For this example we use time-varying sparse 3-column contact matrices from [@Rao:2017aa], representing the HCT-116 cell line. The matrices represent a single chromosome 22 cell sample treated with TAD boundary destroying auxin. The data is then sequenced at 4 time points (20, 40, 60 and 180 minutes) after withdrawal of auxin. Once auxin is withdrawn, the TAD boundaries slowly return. Using `TimeCompare` we can track the return of TADs after withdrawal.

```{r, message = FALSE, warning = FALSE}
#Get the list of contact matrices
data("time_mats")
#Checking format
head(time_mats[[1]])
#These are sparse 3-column matrices
#Running MultiCompare
time_var = TimeCompare(time_mats, resolution = 50000)
#Print head of all regions containing a TAD at one time point
head(time_var$TAD_Bounds)
```

## Running ConsensusTADs

`ConsensusTADs` is an approach for calling TADs across multiple datasets. It uses the median boundary score across n replicates or cell lines to calculate a consensus of TAD boundaries. This effectively filters out noisy TAD boundaries that may appear in only one or a few replicates or cell lines. `ConsensusTADs` takes essentially the same input as `TimeCompare` (a list of contact matrices) and provides consensus TAD scores for each region, summarized across each contact matrix of the genome. It also provides a list of regions with significant TAD scores. These regions can be thought of as consensus TAD boundaries. Using these, we can get a single set of TADs summarized across a set of replicates, conditions or time points. For this example we use the two replicates from [@Rao:2014aa].

```{r, message = FALSE, warning = FALSE}
#Get the rao contact matrices built into the package
data("rao_chr22_prim")
data("rao_chr22_rep")
#Place into list
cont_mats = list(rao_chr22_prim, rao_chr22_rep)
#Running MultiCompare
con_tads = ConsensusTADs(cont_mats, resolution = 50000)
#Print head of all regions containing a TAD at at least one time point
head(con_tads$Consensus)
```

## Effect of matrix type on runtime

The type of matrix input into the algorithm can affect runtimes for the algorithm. $n \times n$ matrices require no conversion and are the fastest. Meanwhile, $n \times (n+3)$ matrices take slightly longer to run due to the need to remove the first 3 columns. Sparse 3-column matrices have the highest runtimes due to the complexity of converting them to an $n \times n$ matrix. The times are summarized below, holding all other parameters constant.

```{r, message = FALSE}
library(microbenchmark)
#Converting to sparse
prim_sparse = HiCcompare::full2sparse(rao_chr22_prim)
rep_sparse = HiCcompare::full2sparse(rao_chr22_rep)
#Converting to nxn+3
#Primary
prim_n_n_3 = data.frame(chr = "chr22",
                        start = as.numeric(colnames(rao_chr22_prim)),
                        end = as.numeric(colnames(rao_chr22_prim))+50000, 
                        rao_chr22_prim)

#Replicate
rep_n_n_3 = data.frame(chr = "chr22", 
                       start = as.numeric(colnames(rao_chr22_rep)),
                       end = as.numeric(colnames(rao_chr22_rep))+50000,
                       rao_chr22_rep)
#Defining each function
#Sparse
sparse = TADCompare(cont_mat1 = prim_sparse, cont_mat2 = rep_sparse, resolution = 50000)
#NxN
n_by_n = TADCompare(cont_mat1 = prim_sparse, cont_mat2 = rep_sparse, resolution = 50000)
#Nx(N+3)
n_by_n_3 = TADCompare(cont_mat1 = prim_n_n_3, cont_mat2 = rep_n_n_3, resolution = 50000)

#Benchmarking different parameters
microbenchmark(
#Sparse
sparse = TADCompare(cont_mat1 = prim_sparse, cont_mat2 = rep_sparse, resolution = 50000),
#NxN
n_by_n = TADCompare(cont_mat1 = rao_chr22_prim, cont_mat2 = rao_chr22_rep, resolution = 50000),
#Nx(N+3)
n_by_n_3 = TADCompare(cont_mat1 = prim_n_n_3, cont_mat2 = rep_n_n_3, resolution = 50000), times = 5, unit = "s"
)
```

The table above shows a summary of runtimes for different types of contact matrices measured in seconds. As we see, TADCompare is extremely fast no matter the parameters. However, sparse matrix inputs will slow down the algorithm. This can become more apparent as the size of the contact matrices increase.

# References
